---
title: "Exponential Smoothing"
author: "Jeremy O'Brien & Juliann McEachern"
date: "September 17, 2019"
output:
  ioslides_presentation:
    theme: cosmo
    highlight: kate
    css: style.css
    widescreen: true
---

```{r, echo=F, comment=F, error=F, warning=F, message=F}
library(kableExtra)
library(formattable)
library(dplyr)
library(fpp2)
library(ggplot2)
```


## Agenda

1.  Overview  
2.  Simple methods  
3.  Trend methods  
4.  Seasonal methods
5.  Taxonomy of Exponential Smoothing Methods
6.  Innovations State Space Models 
7.  Estimation and Model Selection
8.  Forecasting with ETS models

# Overview

## Overview | What's exponential smoothing?

* When making predictions, weight recent observations more than older observations  
* Specifically, decrease weights geometrically / exponentially when moving backward  
* Window function (a.k.a. apodization function or tapering function)  
  + zero-valued outside of some chosen interval  
  + normally symmetric around middle of interval  
  + typically near maximum in the middle and tapering away from there  
* Easily learned and applied (i.e. signal processing community in 1940's)

# Simple Exponential Smoothing 

## Simple Exponential Smoothing |  SES Methods

Time series $y_1,y_2,\dots,y_T$.

**Random walk forecast**

$$
\begin{aligned}
  \hat{y}_{T+h|T} = y_T
\end{aligned}
$$

**Average forecast** 

$$
\begin{aligned}
   \hat{y}_{T+h|T} = \frac1T\sum_{t=1}^T y_t
\end{aligned}
$$



Consider a happy median that privileges recent information...

## Simple Exponential Smoothing |  SES Methods

**Forecast equation**
$$
\begin{aligned}
   \hat{y}_{T+1|T} = \alpha y_T + \alpha(1-\alpha) y_{T-1} + \alpha(1-\alpha)^2 y_{T-2}+ \cdots
\end{aligned}
$$
where $0 \le \alpha \le1$

Weighted moving average with weights that decrease exponentially

* for $y_{T}$ and $\alpha = .2$, the weighting is $.2$
* for $y_{T-1}$, the weighting is $.2 * (1 - .2) = .16$
* for $y_{T-2}$, $.2 * (1 - .2)^2 = .128$
* etc.


## Simple Exponential Smoothing |  Component Form

**Component form: forecast equation**
$$
\begin{align*}
   \hat{y}_{t+h|t} = \ell_{t}
\end{align*}
$$

**Component form: smoothing equation**
$$
\begin{align*}
   \ell_{t} = \alpha y_{t} + (1 - \alpha)\ell_{t-1}
\end{align*}
$$

$\ell_t$ is level / smoothed value of series at time $t$

$\hat{y}_{t+1|t} = \alpha y_t + (1-\alpha) \hat{y}_{t|t-1}$  
Iterate for exponentially weighted moving average form


## Simple Exponential Smoothing |  Weighted Average Form
 
**Weighted average form**
$$
\begin{align*}
   \hat{y}_{T+1|T}=\sum_{j=0}^{T-1} \alpha(1-\alpha)^j y_{T-j}+(1-\alpha)^T \ell_{0}
\end{align*}
$$


## Simple Exponential Smoothing | Optimization
 
To choose value for $\alpha$ and $\ell_0$, minimize SSE:
$$
\begin{aligned}
    \text{SSE}=\sum_{t=1}^T(y_t - \hat{y}_{t|t-1})^2.
\end{aligned}
$$

Unlike regression, no closed-form solution so use numerical optimization


## Simple Exponential Smoothing |  Example: Oil production
 
```{r ses1, eval=F, echo=TRUE, cache=TRUE}
oildata <- window(oil, start=1996)
fc <- ses(oildata, h=5)
autoplot(fc) +
  autolayer(fitted(fc), series="Fitted") +
  ylab("Oil (millions of tonnes)") + xlab("Year")
```

## Simple Exponential Smoothing |  Example: Oil production {.flexbox .vcenter}

```{r ses2, echo=F, cache=TRUE}
oildata <- window(oil, start=1996)
fc <- ses(oildata, h=5)
autoplot(fc) +
  autolayer(fitted(fc), series="Fitted") +
  ylab("Oil (millions of tonnes)") + xlab("Year")
```

# Trend Methods

## Trend Methods | Holt's Linear Trend
 
**Component form: forecast**
$$
\begin{align*}
   \hat{y}_{t+h|t} &= \ell_{t+h}b_{t}
\end{align*}
$$

**Component form: level**
$$
\begin{align*}
    \ell_{t} &= \alpha y_{t} + (1 - \alpha)(\ell_{t-1} + b_{t-1})
\end{align*}
$$

**Component form: trend**
$$
\begin{align*}
   b_{t} &= \beta^*(\ell_{t} - \ell_{t-1}) + (1 -\beta^*)b_{t-1}
\end{align*}
$$

## Trend Methods | Holt's Linear Trend (cont'd)

* Two smoothing parameters $\alpha$ and $\beta^*$ $(0\le\alpha,\beta^*\le1)$  
* $\ell_t$ level: weighted average between $y_t$ and one-step ahead forecast for time $t$, $(\ell_{t-1} + b_{t-1} = \hat{y}_{t|t-1})$  
* $b_t$ slope: weighted average of $(\ell_{t} - \ell_{t-1})$ and $b_{t-1}$, current and previous estimate of slope  
* Select $\alpha, \beta^*, \ell_0, b_0$ to minimise SSE


## Trend Methods | Example: Holt's Method
 

```{r, fig.height=3.6, echo=TRUE}
window(ausair, start=1990, end=2004) %>%
  holt(h=5, PI=FALSE) %>%
  autoplot()
```


## Trend Methods | Damped Trend Method
 
**Component form**
$$
\begin{align*}
  \hat{y}_{t+h|t} &= \ell_{t} + (\phi+\phi^2 + \dots + \phi^{h})b_{t} \\
  \ell_{t} &= \alpha y_{t} + (1 - \alpha)(\ell_{t-1} + \phi b_{t-1})\\
  b_{t} &= \beta^*(\ell_{t} - \ell_{t-1}) + (1 -\beta^*)\phi b_{t-1}
\end{align*}
$$

* Damping parameter: $0<\phi<1$
* If $\phi=1$, identical to Holt's linear trend
* As $h\rightarrow\infty$, $\hat{y}_{T+h|T}\rightarrow \ell_T+\phi b_T/(1-\phi)$
* Short-run forecasts trended, long-run forecasts constant


# Seasonal Methods

## Seasonal Methods | Holt-Winters Additive Method
 
Holt's method extended to capture seasonality

**Component form**
$$
\begin{align*}
   \hat{y}_{t+h|t} &= \ell_{t} + hb _{t} + s_{t+h-m(k+1)} \\
   \ell_{t} &= \alpha(y_{t} - s_{t-m}) + (1 - \alpha)(\ell_{t-1} + b_{t-1})\\
   b_{t} &= \beta^*(\ell_{t} - \ell_{t-1}) + (1 - \beta^*)b_{t-1}\\
   s_{t} &= \gamma (y_{t}-\ell_{t-1}-b_{t-1}) + (1-\gamma)s_{t-m},
\end{align*}
$$

* $k$ is integer part of $(h-1)/m$, ensuring final year estimates in forecast
* Parameters: $0\le \alpha\le 1$, $0\le \beta^*\le 1$, $0\le \gamma\le 1-\alpha$ and $m$ for period of seasonality (e.g. $m=4$ for quarterly data)


## Seasonal Methods | Holt-Winters Additive Method
 
* Seasonal component usually expressed as $s_{t} = \gamma^* (y_{t}-\ell_{t}) + (1-\gamma^*)s_{t-m}$
* Substitute for $\ell_t$: $s_{t} = \gamma^* (1-\alpha) (y_{t}-\ell_{t-1}-b_{t-1}) + [1-\gamma^*(1-\alpha)]s_{t-m}$
* Set $\gamma=\gamma^*(1-\alpha)$
* Usual parameter restriction is $0\le\gamma^*\le1$ or $0\le\gamma\le(1-\alpha)$


## Seasonal Methods | Holt-Winters Multiplicative Method
 
When seasonal variations change in proportion to level of series

**Component form**

$$
\begin{align*}
   \hat{y}_{t+h|t} &= (\ell_{t} + hb_{t})s_{t+h-m(k+1)}\\
   \ell_{t} &= \alpha \frac{y_{t}}{s_{t-m}} + (1 - \alpha)(\ell_{t-1} + b_{t-1})\\
   b_{t} &= \beta^*(\ell_{t}-\ell_{t-1}) + (1 - \beta^*)b_{t-1}\\
   s_{t} &= \gamma \frac{y_{t}}{(\ell_{t-1} + b_{t-1})} + (1 - \gamma)s_{t-m}
\end{align*}
$$

  * For additive method, $s_t$ is in absolute terms: within each year $\sum_i s_i \approx 0$
  * For multiplicative method, $s_t$ is in relative terms: within each year $\sum_i s_i \approx m$


## Seasonal Methods | Example: Visitor Nights
 
```{r 7-HW, echo=TRUE, eval=F}
aust <- window(austourists,start=2005)
fit1 <- hw(aust,seasonal="additive")
fit2 <- hw(aust,seasonal="multiplicative")
```

```{r, eval=F}
tmp <- cbind(Data=aust,
  "HW additive forecasts" = fit1[["mean"]],
  "HW multiplicative forecasts" = fit2[["mean"]])

autoplot(tmp) + xlab("Year") +
  ylab("International visitor night in Australia (millions)") +
  scale_color_manual(name="",
    values=c('#000000','#1b9e77','#d95f02'),
    breaks=c("Data","HW additive forecasts","HW multiplicative forecasts"))
```

## Seasonal Methods | Example: Visitor Nights {.flexbox .vcenter}
 
```{r, echo=TRUE, echo=F}
aust <- window(austourists,start=2005)
fit1 <- hw(aust,seasonal="additive")
fit2 <- hw(aust,seasonal="multiplicative")
```

```{r, fig.height=3.2, echo=F}
tmp <- cbind(Data=aust,
  "HW additive forecasts" = fit1[["mean"]],
  "HW multiplicative forecasts" = fit2[["mean"]])

autoplot(tmp) + xlab("Year") +
  ylab("International visitor night in Australia (millions)") +
  scale_color_manual(name="",
    values=c('#000000','#1b9e77','#d95f02'),
    breaks=c("Data","HW additive forecasts","HW multiplicative forecasts"))
```


## Seasonal Methods | Example: Visitor nights estimated components
 
```{r fig-7-LevelTrendSeas1, eval=F}
addstates <- fit1$model$states[,1:3]
multstates <- fit2$model$states[,1:3]
colnames(addstates) <- colnames(multstates) <-
  c("level","slope","season")
p1 <- autoplot(addstates, facets=TRUE) + xlab("Year") +
  ylab("") + ggtitle("Additive states")
p2 <- autoplot(multstates, facets=TRUE) + xlab("Year") +
  ylab("") + ggtitle("Multiplicative states")
gridExtra::grid.arrange(p1,p2,ncol=2)
```

## Seasonal Methods | Example: Vistor nights estimated components {.flexbox .vcenter}
 
```{r fig-7-LevelTrendSeas2, echo=F}
addstates <- fit1$model$states[,1:3]
multstates <- fit2$model$states[,1:3]
colnames(addstates) <- colnames(multstates) <-
  c("level","slope","season")
p1 <- autoplot(addstates, facets=TRUE) + xlab("Year") +
  ylab("") + ggtitle("Additive states")
p2 <- autoplot(multstates, facets=TRUE) + xlab("Year") +
  ylab("") + ggtitle("Multiplicative states")
gridExtra::grid.arrange(p1,p2,ncol=2)
```


## Seasonal Methods | Holt-Winters Damped Method
 
$$
\begin{align*}
   \hat{y}_{t+h|t} &= [\ell_{t} + (\phi+\phi^2 + \dots + \phi^{h})b_{t}]s_{t+h-m(k+1)} \\
   \ell_{t} &= \alpha(y_{t} / s_{t-m}) + (1 - \alpha)(\ell_{t-1} + \phi b_{t-1})\\
   b_{t} &= \beta^*(\ell_{t} - \ell_{t-1}) + (1 - \beta^*)\phi b_{t-1}       \\
   s_{t} &= \gamma \frac{y_{t}}{(\ell_{t-1} + \phi b_{t-1})} + (1 - \gamma)s_{t-m}
\end{align*}
$$

Often single most accurate forecasting method for seasonal data


## Seasonal Methods | Example: Damped trend

```{r fig-8-DampedTrend1, eval=F}
air <- window(ausair, start = 1990, end = 2004)
fit3 <- holt(air, h = 15)
fit4 <- holt(air, damped = TRUE, phi = .9, h = 15)
autoplot(air) +
  autolayer(fit3, series = 'Holt\'s method', PI = FALSE) +
  autolayer(fit4, series = 'Dampled Holt\'s method', PI = FALSE) +
  ggtitle('Forecasts from Holt\'s method') +
  xlab('Year') +
  ylab('Air passengers in Australia (millions)') +
  guides(color = guide_legend(title = 'Forecast'))
```

## Seasonal Methods | Example: Damped trend {.flexbox .vcenter}

```{r fig-8-DampedTrend2, echo=F}
air <- window(ausair, start = 1990, end = 2004)
fit3 <- holt(air, h = 15)
fit4 <- holt(air, damped = TRUE, phi = .9, h = 15)
autoplot(air) +
  autolayer(fit3, series = 'Holt\'s method', PI = FALSE) +
  autolayer(fit4, series = 'Dampled Holt\'s method', PI = FALSE) +
  ggtitle('Forecasts from Holt\'s method') +
  xlab('Year') +
  ylab('Air passengers in Australia (millions)') +
  guides(color = guide_legend(title = 'Forecast'))
```


# Taxonomy of Exponential Smoothing Methods 

## Taxonomy | Exponential Smoothing Methods

<table class="table table-striped" style="margin-left: auto; margin-right: auto;">
 <thead>
<tr>
<th style="border-bottom:hidden" colspan="1"></th>
<th style="border-bottom:hidden; padding-bottom:0; padding-left:3px;padding-right:3px;text-align: center; " colspan="3"><div style="padding-bottom: 5px;">SEASONAL COMPONENT</div></th>
</tr>
  <tr>
   <th style="text-align:left; text-align: center" colspan="1"> 
   <div style="padding-bottom: 5px;"><strong>TREND COMPONENT</strong></div></th>
   <th style="text-align:center;width: 3cm;"> None <br> <font color=#ffd690>(*N*)</font> </th>
   <th style="text-align:center;width: 3cm;"> Additive <br> <font color=#ec990b>(*A*)</font> </th>
   <th style="text-align:center; width: 3cm;"> Multiplicative <br> <font color=#c08726>(*M*)</font></th>
  </tr>
 </thead>
<tbody>
  <tr>
   <th style="text-align:center;width: 5cm; "> None <br> 
   <font color=#76b7d7>(*N*)</font> </div></th>
   <td style="text-align:center;"> 
   (<font color=#76b7d7>N</font>,
   <font color=#ffd690>N</font>) 
   </td>
   <td style="text-align:center;"> 
   (<font color=#76b7d7>N</font>,
   <font color=#ec990b>A</font>) 
   </td>
   <td style="text-align:center;"> 
   (<font color=#76b7d7>N</font>, 
   <font color=#c08726>M</font>)
   </td>
  </tr>
  <tr>
   <th style="text-align:center;width: 5cm; "> Additive <br> 
   <font color=#137db3>(*A*)</font> </td>
   <td style="text-align:center;"> 
   (<font color=#137db3>A</font>,
   <font color=#ffd690>N</font>)
   </td>
   <td style="text-align:center;"> 
   (<font color=#137db3>A</font>,
   <font color=#ec990b>A</font>)
   </td>
   <td style="text-align:center;"> 
   (<font color=#137db3>A</font>, 
   <font color=#c08726>M</font>)
   </td>
  </tr>
  <tr>
   <th style="text-align:center;width: 5cm; "> Additive Damped <br> 
   <font color=#1f5976>(*A$_d$*)</font></td>
   <td style="text-align:center;"> 
   (<font color=#1f5976>A$_d$</font>, 
   <font color=#ffd690>N</font>)
   </td>
   <td style="text-align:center;"> 
   (<font color=#1f5976>A$_d$</font>, 
   <font color=#ec990b>A</font>) 
   </td>
   <td style="text-align:center;"> 
   (<font color=#1f5976>A$_d$</font>, 
   <font color=#c08726>M</font>)
   </td>
  </tr>
</tbody>
</table>

## Taxonomy | Variations Explained 

Short hand representation of methods: 

*  (<font color=#76b7d7>N</font>, <font color=#ffd690>N</font>):  Simple exponential smoothing
*  (<font color=#137db3>A</font>, <font color=#ffd690>N</font>):  Holt’s linear method 
*  (<font color=#1f5976>A$_d$</font>, <font color=#ffd690>N</font>):  Additive damped trend method 
*  (<font color=#137db3>A</font>, <font color=#ec990b>A</font>):  Additive Holt-Winters’ method 
*  (<font color=#137db3>A</font>, <font color=#c08726>M</font>):  Multiplicative Holt-Winters’ method 
*  (<font color=#1f5976>A$_d$</font>, <font color=#c08726>M</font>):  Holt-Winters’ damped method 

## Taxonomy | R Functions {.smaller}
 
Simple Exponential Smoothing: No Trend  
>  `ses(y)`

Holt's Method: Linear Trend   
>  `holt(y)`

Damped trend Method.  
>  `holt(y, damped=TRUE)`

Holt-Winters methods  
>  `hw(y, damped=TRUE, seasonal="additive")`  
>  `hw(y, damped=FALSE, seasonal="additive")`  
>  `hw(y, damped=TRUE, seasonal="multiplicative")`  
>  `hw(y, damped=FALSE, seasonal="multiplicative")`

# Innovations State Space Models

## Innovations State Space Models | Exponential Smoothing Models vs. Methods

1.  **Methods**: 
    Apply algorithms to obtain point forecasts
2.  **Models**: 
    Apply stochastic generating process to forecast an entire  distribution.
    +  Generate point forcasts *and* forecast intervals
    +  Beneficial for model selection 

## Innovations State Space Models | Overview

1.  Each model contains two equations:   
    +  **Observation** equation for observed data components.  
    +  **State** equation for unobserved changes in level, trend, and seasonal components.  
    
2.  Each method contains two models distinguished by either additive or multiplicative errors.  

3.  **E**rror, **T**rend, **S**easonal (**ETS**) labeling is used to differentiate models, methods, and type.

## Innovations State Space Models | ETS Structure

18 variations of ETS models exist and are classified as follows:  

1.  Error: $\{$<font color=#9dd182>$A$</font> | 
              <font color=#659b4a>$M$</font>$\}$
    +  Additive Errors
    +  Multipicative Errors

2.  Trend: $\{$<font color=#76b7d7>$N$</font> | 
            <font color=#137db3>$A$</font> | 
            <font color=#1f5976>$A_d$</font>$\}$ 

3.  Seasonal: $\{$<font color=#ffd690>$N$</font> | 
              <font color=#ec990b>$A$</font> | 
              <font color=#c08726>$M$</font>$\}$
              
## Innovations State Space Models | ETS Model Classification Examples
*  $($<font color=#9dd182>$A$</font>, 
<font color=#76b7d7>$N$</font>, 
<font color=#ffd690>$N$</font>$)$:  Simple exponential smoothing; additive errors
*  $($<font color=#659b4a>$M$</font>,
    <font color=#76b7d7>$N$</font>, 
    <font color=#ffd690>$N$</font>$)$: Simple exponential smoothing; multiplicative errors  
*  $($<font color=#9dd182>$A$</font>,
    <font color=#137db3>$A$</font>, 
    <font color=#ffd690>$N$</font>$)$:  Holt's linear method; additive errors 
*  $($<font color=#659b4a>$M$</font>,
      <font color=#137db3>$A$</font>, 
      <font color=#ffd690>$N$</font>$)$: Holt’s linear method; multiplicative errors 
*  $($<font color=#659b4a>$M$</font>,
      <font color=#137db3>$A$</font>, 
      <font color=#c08726>$M$</font>$)$:  Multiplicative Holt-Winters’ method; multiplicative errors

# Estimation and Model Selection

##  Estimation and Model Selection | Estimating ETS models

*  Smoothing parameters $\alpha$, $\beta$, $\gamma$ and $\phi$, and the initial states $\ell_0$, $b_0$, $s_0,s_{-1},\dots,s_{-m+1}$ are estimated by maximising the "likelihood" = the probability of the data arising from the specified model.
*  For models with additive errors equivalent to minimising SSE.
*  For models with multiplicative errors, **not** equivalent to minimising SSE.
*  We will estimate models with the `ets()` function in the forecast package.

##  Estimation and Model Selection | Selecting ETS models

**Akaike’s Information Criterion (AIC):**

$$
\text{AIC} = -2\log(\text{L}) + 2k
$$

**Corrected Akaike’s Information Criterion (AIC$_c$):**

$$
\text{AIC}_{\text{c}} = \text{AIC} + \frac{2(k+1)(k+2)}{T-k}
$$

**Schwarz’s Bayesian Information Criterion (BIC):**

$$
\text{BIC} = \text{AIC} + k(\log(T)-2)
$$



##  Estimation and Model Selection | Automatic Forecasting: Tourist Example

Automatic forecasting using the `ets()` function: 
*  Automatically selects models when arguments are set to default.
*  Produces forecasts using best method.
*  Obtain forecast intervals using underlying state spacemodel.

Tourist example from text: 

```{r, eval=F}
aust <- window(austourists, start=2005) 
fit <- ets(aust) 
summary(fit)
```

##  Estimation and Model Selection | Automatic Forecasting: Tourist Example {.smaller}

```{r, echo=F, tidy=F, out.width="30%", out.height="20%"}
aust <- window(austourists, start=2005) 
fit <- ets(aust) 
summary(fit)
```

# Forecasting with ETS models

##  Forecasting with ETS models

## Sources

1. https://github.com/robjhyndman/ETC3550Slides/raw/master/7-exponentialsmoothing.pdf
2. https://en.wikipedia.org/wiki/Exponential_smoothing
3. https://machinelearningmastery.com/exponential-smoothing-for-time-series-forecasting-in-python/
